Einstein's gravity from a polynomial affine model

### 2016-04-12 AffGR-PRD.tex
Improved conclusions, unfinished

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between the models does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is torsion-free. Notice that the vanishing torsion limit---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$---cannot be taken at the action level, but in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements. It can be realized as follows, since only the terms of the action linear in these fields will be relevant, one can consider the effective action linear in the torsion's fields, i.e.,
\begin{equation}
  %\label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\lambda\mu}{}^{\lambda}{}_\nu \nabla_\rho %T^{\nu,\rho\sigma}
  + C_2 \, R_{\mu\rho}{}^{\lambda}{}_\lambda \nabla_\nu \Big) T^{\nu,\mu\rho} ,
\end{equation}
and the only nontrivial field equation after the limit will be the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
\end{equation}
with $\kappa$ a constant related with the original couplings of the model. We assume that the connection is compatible with volume form, i.e., it is equi-affine~\cite{nomizu1994affine,MO-Bryant02}.  The equi-affine condition assures that the Ricci tensor of the connection is symmetric, and the contraction of the last indices vanishes, thus the second term in the field equation is absent and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

@@

The gravitational model described by the action in Eq.~\eqref{SKY} is called Stephenson--Kilmister--Yang (or SKY for short). However, the standard interpretation of this model requires a metric tensor. A quick analysis of the field equations \`a la Palatini shows that Eqs.~\eqref{SimpleEOM} and~\eqref{harm-curv} are obtained only from the variation with respect to the connection. Albeit this effective theory has been widely studied, the extra field equation (for the metric field) strongly constraints the solutions of the gravitational Yang--Mills~\footnote{We thank J.~Zanelli for sharing his personal communications with C.~N.~Yang, and clarify this aspect of the effective model.}. However, that complication does not affect our affine interpretation of the theory, where the metric is absent.

@@

Summarizing, we have shown that the  \emph{purely affine gravity} constructed in Ref.~\cite{Skirzewski:2014eta} possesses a well-defined torsionless limit, which generalizes the equations of motion  from General Relativity, and its field equations are equivalent to those obtained from a Yang--Mills theory for the spacetime connection, or Stephenson--Kilmister--Yang theory. Consequently, all gravitational effects described by standard General Relativity are contained in our model, even though the torsion field is kept to zero. Regarding this aspect, we would like to emphazise that there is enough liberty to start considering new cosmological effects coming from both nonvanishing torsion and nonmetric connections with vanishing torsion, beside the results reported in Ref.~\cite{Chen:2013kia,Chen:2013ota}.
Moreover, our gravitational model coupled to a scalar field through the \emph{tensor density} $\mathfrak{g}^{\mu\nu}$ ---in the vanishing torsion consideration---, generalizes GR minimally coupled to a massless scalar field. In a forthcoming paper, we will consider more general matter content (unrelated to the work in Ref.~\cite{Cook:2008mx}), and some properties of the field equations which differ from those of General Relativity.

%%%%%%%%% ACKNOWLEGMENTS %%%%%%%%%
\begin{acknowledgments}
  We thank to N.~Pantoja, A.~Melfo and R.~L.~Bryant for their helpful discussions and inspiring comments, to J. Zanelli for his suggestions on the physical insight into the problem and careful but critical review of the manuscript.
  This work was partially supported by CONICYT (Chile) under project No. 79140040.
\end{acknowledgments}

>>>>

%%\hl{A comparison between our affine model and GR requires the imposition of vanishing torsion (JZ: Why?).}
Even though a comparison between the models does not require the imposition of vanishing torsion, for the sake of simplicity, at this stage we shall focus on a sector of the theory in which the connection is torsion-free. Notice that the vanishing torsion limit---equivalent to take $T^{\lambda,\mu\nu} \to 0$ and $A_\mu \to 0$---cannot be taken at the action level, but in the field equations. Interestingly, the field equations from Eq.~\eqref{4dfull} can be consistently \emph{truncated} under such requirements. It can be realized as follows, since only the terms of the action linear in these fields will be relevant, one can consider the effective action linear in the torsion's fields, i.e.,
\begin{equation}
  \label{eff-action}
  S_{\text{eff}} = \int\dn{4}{x} \Big( C_1\, R_{\lambda\mu}{}^{\lambda}{}_\nu \nabla_\rho %T^{\nu,\rho\sigma}
  + C_2 \, R_{\mu\rho}{}^{\lambda}{}_\lambda \nabla_\nu \Big) T^{\nu,\mu\rho} ,
\end{equation}
and the only nontrivial field equation after the limit will be the one for the Curtright, $T^{\nu,\mu\rho}$,
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} + \kappa \nab{\nu} R_{\mu\rho}{}^\lambda{}_\lambda = 0,
  \label{almostSimpleEOM}
\end{equation}
with $\kappa$ a constant related with the original couplings of the model. We assume that the connection is compatible with volume form, i.e., it is equi-affine~\cite{nomizu1994affine,MO-Bryant02}.  The equi-affine condition assures that the Ricci tensor of the connection is symmetric, and the contraction of the last indices vanishes, thus the second term in the field equation is absent and the gravitational equations are
\begin{equation}
  \nab{[\rho} R_{\mu]\nu} = 0.
  \label{SimpleEOM}
\end{equation}

@@

The gravitational model described by the action in Eq.~\eqref{SKY} is called Stephenson--Kilmister--Yang (or SKY for short). However, the standard interpretation of this model requires a metric tensor. A quick analysis of the field equations \`a la Palatini shows that Eqs.~\eqref{SimpleEOM} and~\eqref{harm-curv} are obtained only from the variation with respect to the connection. Albeit this effective theory has been widely studied, the extra field equation (for the metric field) strongly constraints the solutions of the gravitational Yang--Mills~\cite{JZcomm}. However, that complication does not affect our affine interpretation of the theory, where the metric is absent.

@@

In this paper we have shown that the \emph{purely affine gravity} presented in Ref.~\cite{Skirzewski:2014eta} is the most general (polynomial) model for an affine theory of gravity, see the Appendix~\ref{sec:ld}, consistent with invariance under diffeomorphisms. The action in Eq.~\eqref{4dfull}, is parametrized by a nineteen-dimensional moduli space---after the boundary and topological terms have been dropped---, which in some sense made us wonder about a possible \emph{duality} between this model and the one proposed in Ref.~\cite{Pagani:2015ema}, even though the last one possesses a metric field.
%% (boundary condition, topological terms, ...)

The field equations obtained from the action~\eqref{4dfull} are a set of very complex coupled partial differential equations. However, they are consistent with the limit of $T^{\lambda,\mu\nu}$ and $A_\mu$ going to zero. This consistency is what is usually called a \emph{consistent truncation} of the theory. In this paper we have restrict ourselves to such a truncation of our model, which constrained us to the torsion-less sector of the moduli space. Additionally, this truncation leaves only one nontrivial field equation, Eq.~\eqref{SimpleEOM}.
%%, which can be obtained from a simple ``effective'' theory whose action is Eq.~\eqref{eff-action}.
Moreover, by considering torsion-free equi-affine connections---those preserving a volume element---or a Levi-Civita connection---in the cases where the manifold is metric---, the remaining field equation simplifies even further to Eq.~\eqref{SimpleEOM}.

The equation~\eqref{SimpleEOM} is a generalization of the field equations in General Relativity, in the sense that, any Einstein manifold (whose Ricci tensor is proportional to the metric) satisfy the parallel Ricci condition,
\begin{equation}
  \nab{\lambda} R_{\mu\nu} = 0.
\end{equation}
Although all Einstein manifolds satisfy the parallel Ricci condition, the contrary does not hold. Therefore, in Appendix~\ref{Known} we present a summary of known (metric) solutions to Eq.~\eqref{SimpleEOM}. In addition, in Appendix~\ref{Schw} we prove that an ansatz metric with spherical symmetry solves the field equation only if it is static. This is a Birkhoff-like theorem for our model, however, the non-linearity of the equations does not assure the uniqueness of the solution. Nonetheless, certain equations become linear if the functions $A$ and $B$ in Eq.~\eqref{ABmetric} are equal. Therefore, the unique solution is the $\Lambda$-Schwarzschild solution.

We also pointed that our field equation~\eqref{SimpleEOM} can be obtained from an effective action, known as Stephenson--Kilmister--Yang (SKY) theory~\cite{stephenson1958quadratic,kilmister1961use,Yang1974}, which is built up in the spirit of a Yang--Mills theory for gravity. Despite the SKY theory has been criticized by many people, including C.~N.~Yang himself~\cite{JZcomm}, because its field equations are higher order differential equations for the metric field, and also because the field equation for the metric is so restrictive that the Schwarzschild metric is not a solution, in our model the fundamental gravitational field is the connection. The field equation~\eqref{SimpleEOM} is a second order partial differential equation for the connection. Although we have presented a series of metric solutions of the effective field equations (restricted to the torsion-free sector of the theory), the correct way to proceed is to propose a connection ansatz compatible with the symmetries of the problem. Using this procedure, we expect to find a non-metric solution to the field equation~\cite{OCF-future2}.

We would like to highlight that, since all gravitational effects described by standard General Relativity are contained in our model, particularly in the sector of metric compatible, torsion-free connections, there is enough liberty to start considering new cosmological effects coming from both nonvanishing torsion and non-metric connections with vanishing torsion, beside the results reported in Ref.~\cite{Chen:2013kia,Chen:2013ota}.

As a first step toward the coupling of matter with our polynomial affine gravity, we showed that there is a construct which might play the role of an inverse metric density. Notice that such a tensor density \emph{does not} necessarily satisfy a non-degeneracy condition as required for a metric, but even though it is an interesting question to ask under what considerations does this construct satisfy the properties of a metric. Using this tensor density, we proposed a \emph{kinetic} term for a scalar field (under diffeomorphisms), and show that in generically the effective field equations for gravity are equivalent to those of General Relativity, when we restrict ourselves to the torsion-free sector. Despite the fact that in this sector the field equation for the scalar field collapses, and cannot be obtained from the variation of the action, using the algorithm proposed in Ref.~\cite{Bekenstein:2014uwa}, the usual scalar field equation is recovered in virtue of the second Bianchi identity.

Finally, we analyzed the second order perturbation of the action in Eq.~\eqref{4dfull}, around a maximally symmetric spacetime.

{\huge{\textcolor{red}{Working Here!}}}

%% Summarizing, we have shown that the \emph{purely affine gravity} presented in Ref.~\cite{Skirzewski:2014eta} possesses a well-defined torsionless limit, which generalizes the equations of motion  from General Relativity, and its field equations are equivalent to those obtained from a Yang--Mills theory for the spacetime connection, or Stephenson--Kilmister--Yang theory. Consequently, all gravitational effects described by standard General Relativity are contained in our model, even though the torsion field is kept to zero. Regarding this aspect, we would like to emphazise that there is enough liberty to start considering new cosmological effects coming from both nonvanishing torsion and nonmetric connections with vanishing torsion, beside the results reported in Ref.~\cite{Chen:2013kia,Chen:2013ota}.
%% Moreover, our gravitational model coupled to a scalar field through the \emph{tensor density} $\mathfrak{g}^{\mu\nu}$ ---in the vanishing torsion consideration---, generalizes GR minimally coupled to a massless scalar field. In a forthcoming paper, we will consider more general matter content (unrelated to the work in Ref.~\cite{Cook:2008mx}), and some properties of the field equations which differ from those of General Relativity.

%%%%%%%%% ACKNOWLEGMENTS %%%%%%%%%
\begin{acknowledgments}
  We thank to N.~Pantoja, A.~Melfo, M.~Blagojevi\'c, B.~Cvetkovi\'c, V.~Sharma, O.~Orellana and R.~L.~Bryant for their helpful discussions and inspiring comments, to J.~Zanelli for his suggestions on the physical insight into the problem and careful but critical review of the manuscript.
  This work was partially supported by CONICYT (Chile) under project No. 79140040.
\end{acknowledgments}
